---
title: "Association Rule Mining using Groceries Dataset"
author: "Sahitya Sundar Raj Vijayanagar"
date: "11/08/2021"
output: pdf_document
---

``` {r setup,include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

require(tidyverse)
require(arules)
require(arulesViz)
require(tidyr)
require(RColorBrewer)
```

Load the dataset from 'groceries.txt' as transactions of format basket. As shown, there are 2159 transactions with just one item in the basket, 1643 transactions with two items in the basket, and so on.
``` {r a}
## Load the dataset
groceries = read.transactions("groceries.txt",format='basket',sep=',')

## Summarise the dataset
str(groceries)
summary(groceries)

```


Create the Item frequency plot to indicate the frequencies for different bought items. As shown, whole milk is the most frequently bought grocery item, followed by other vegetables and rolls/buns.
``` {r b}
itemFrequencyPlot(groceries,topN=20,type="absolute",col=brewer.pal(8,'Pastel2'), main="Absolute Item Frequency Plot")

```

Creating association mining rules by randomly selecting minimum support as 0.001, confidence as 0.8, and max length=10, results in 410 rules.
``` {r c}
# Minimum Support as 0.001, confidence as 0.8.
association_rules <- apriori(groceries, parameter = list(supp=0.001, conf=0.8,maxlen=10))

summary(association_rules)
inspect(association_rules[1:10])

```
Looking at the above result, 100% of customers who bought {rice,sugar} also bought whole milk. Similarly, 90.48% of customers who bought {liquor, red/blush wine} also bought bottled beer. 

Trying more stricter rules with conf=0.9 and shorter rules with maxlen=3, results in only 10 rules as shown below:
``` {r d}
shorter.association.rules <- apriori(groceries, parameter = list(supp=0.001, conf=0.9,maxlen=3))
summary(shorter.association.rules)
inspect(shorter.association.rules[1:10])

```

Considering the 410 association rules created by minimum support as 0.001, confidence as 0.8, and max length=10, the next step included removing subsets of larger rules, which resulted in a total of 319 rules.
``` {r e}
subset_rules <- which(colSums(is.subset(association_rules, association_rules)) > 1) # get subset rules in vector
length(subset_rules) 

subset_association_rules <- association_rules[-subset_rules] # remove subset rules.
summary(subset_association_rules)
inspect(subset_association_rules[1:10])

plot(subset_association_rules)
plot(subset_association_rules,method="two-key plot")

```

Looking at the most bought item in the item frequency list, i.e.m, 'whole milk', it is possible to find the items that are most likely to be bought before buying whole milk by using appearance. Further,using conf=1, will indicate the items where 100% of customers bought whole milk, after buying these items.

``` {r f}
wholemilk_association_rules <- apriori(groceries, parameter = list(supp=0.001, conf=1),appearance = list(default="lhs",rhs="whole milk"))

wholemilk_association_rules <- sort(wholemilk_association_rules, by='confidence',decreasing=TRUE)

summary(wholemilk_association_rules)
inspect(wholemilk_association_rules[1:10])

plot(plot(wholemilk_association_rules, method='graph'))
```
As shown above, 100% of customers who bought {rice,sugar}, {canned fish,hygiene articles}, {butter,rice,root vegetables}, etc. have bought whole milk, Similarly, we find 20 such antecedents for whole milk. 

Sorting rules based on confidence:
``` {r g}

top_rules_conf <- sort(subset_association_rules, by='confidence', decreasing=TRUE)
inspect(head(top_rules_conf)) #High-confidence rules

top10subsets_conf <- head(subset_association_rules, n = 10, by = "confidence")
plot(top10subsets_conf, method = "graph")
```
A confidence of 1 indicates that whenever the items on the antecedent are bought, 100% of customers bought the item(s) on the consequent. 


Sorting rules based on lift:
``` {r h}
top_rules_lift <- sort(subset_association_rules, by='lift', decreasing=TRUE)
inspect(head(top_rules_lift)) #High lift rules

top10subsets_conf <- head(subset_association_rules, n = 10, by = "lift")
plot(top10subsets_conf, method = "graph")

```
A rule with lift of 11.23 for {liquore,red/blush wine}->{bottled beer} indicates that the items in the antecedent and consequent are ~11 times more likely to be bought together than bought individually.
